---
title: "`futureheatwaves` package vignette"
author: "G. Brooke Anderson and Colin Eason"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## What the package does

The `futureheatwaves` package can process climate data files that you have stored locally to generate a list of heatwaves in each projection as well as characteristics of each heatwave (e.g., length, intensity, timing in the year). 

## A basic example of using the package

For these functions to work correctly, you must have the climate projections saved in a specific structure locally on your computer. 

Define the following variables: 
- `dataFolder`: Pathway to the directory that contains your climate projections (Note: these will need to be saved within this directory in a specific way, as explained later in this vignette.)
- `citycsv`: Pathway to a comma-separated file with latitude and longitude coordinates for each community for which you want to generate heatwave files. (Note: This file will need to have specific column names for latitude and longitude, as explained later.)

For example, I can specify `dataFolder` and `citycsv` as: 

```{r}
dataFolder <- "~/Downloads/sample/cmip5/" # Note-- final `/` is required for this to work
citycsv <- "inst/cities.csv"
```

You will also need to specify where the output files should be locally saved. Save this as the object `out`: 

```{r}
out <- "~/Downloads/sample/results"
```

